---
title: "Exploratory Data Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  
---
```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, fig.align = "center")

library(tidyverse)
library(haven)
library(lubridate)
library(here)
library(patchwork)
library(epitools)
library(ggcorrplot)
library(patchwork)
library(plotly)

source(here("source", "load_clean_brfss.R"))
source(here("source", "filter_brfss_data.R"))
source(here("source", "standardize_brfss_variable.R"))
source(here("source", "prop_test_functions.R"))
source(here("source", "run_fishers_test.R"))

brfss_data = load_clean_brfss(here("data", "brfss_clean_2017_2024.csv.zip"))
state_fips = read_csv(here("data", "legal_sports_report", "state_fips.csv"))

state_legal_dates = read_csv(here("data", "legal_sports_report", 'state_legalization_dates.csv'))
sb_rev_by_month = read_csv(here("data", "legal_sports_report", "sb_rev_by_month.csv"))
sb_rev_by_state_month = read_csv(here("data", "legal_sports_report",'sb_rev_by_state_month.csv'))
sb_rev_by_state = read_csv(here("data", "legal_sports_report", "sb_rev_by_state.csv"))
state_pop = read_csv(here("data", "state_population_census.csv"))
state_handle_channel = read_csv(here("data","legal_sports_report", "state_handle_by_channel.csv"))

years_in_data = unique(year(pull(sb_rev_by_month, month)))

outcome_vars = c("any_physical_health_not_good_days", "any_mental_health_not_good_days", "has_depressive_disorder", "has_binge_drink")
```



## Sports Betting

We begin our analysis by exploring trends in the sports betting industry itself and how these trends align with the growing number of states that legalized sports betting over time. We looked at total handle which is the total amount wagered regardless of the outcome of the bet (as opposed to revenue).

```{r}
get_n_states_legal = function(input_month) {
  n_states_legal = nrow(filter(state_legal_dates, first_start <= input_month))
}

sb_rev_by_month = sb_rev_by_month |> 
  mutate(
    n_states_legal = sapply(month, get_n_states_legal)
  )

sb_rev_by_month = sb_rev_by_month |>
  mutate(
    handle_B = handle / 1e9 
  )

scale_factor = max(pull(sb_rev_by_month, n_states_legal)) / max(pull(sb_rev_by_month, handle_B))

sb_rev_by_month |> 
  ggplot(aes(x = month)) +
  geom_line(aes(y = handle_B, color = "handle"), linewidth = 1.1) +
  geom_line(aes(y = n_states_legal / scale_factor, color = "legal"), linewidth = 1.1) +
  
  scale_y_continuous(
    name = "Handle (in USD, Billions)",
    labels = scales::comma, 
    
    sec.axis = sec_axis(~ . * scale_factor, name = "States Legal", labels = scales::comma)
  ) +
  
  scale_color_manual(
    name = NULL,
    values = c("handle" = "darkgreen", "legal" = "cornflowerblue"),
    labels = c("handle" = "Total Handle", "legal" = "n States Legal")
  ) +
  
  labs(
    title = "Total Sports Handle and Number of States with Legal Sports Betting",
    x = "Date"
  ) +
  theme_minimal()
```

The plot shows handle going up with the number of states that have legalized sports betting. It also clearly shows strong cyclical trends in total handle, suggesting that the volume of betting is highly sensitive to the sports calendar.


### Understanding Spikes in Sports Betting Handle

In addition to the BRFSS survey, we looked at sports betting data (Total Handle) to understand the cyclic nature of the betting market that could influence mental health outcomes. The purpose is to determine if the spikes in betting activity align with major sports seasons.

#### NFL Season
```{r}
nfl_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "09", "01", sep = "-")),
    ymd(paste(years_in_data, "01", "01", sep = "-")),
    ymd(paste(years_in_data, "02", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 2 ~ "Solid", 
    TRUE ~ "Dotted" 
    ),
  
  label = case_when(
    month(date) == 9 ~ "Start of Season",
    month(date) == 1 ~ "End of Season",
    month(date) == 2 ~ "Super Bowl"
    )
  ) |> 
  filter(date >= min(sb_rev_by_month$month), date <= max(sb_rev_by_month$month))


sb_rev_by_month |>
  mutate(handle_B = handle / 1e9) |>
  ggplot(aes(x = month, y = handle_B)) +
  geom_line() +
  ylab("Handle (in USD, Billions)") +
  geom_vline(
    data = nfl_event_dates,
    aes(xintercept = date, linetype = type),
    color = "#D50A0A",
    show.legend = TRUE 
  ) +
  scale_linetype_manual(
    name = "NFL Season", 
    values = c("Dotted" = "dotted", "Solid" = "solid"), 
    labels = c("Dotted" = "NFL Regular Season", "Solid" = "Super Bowl") 
  ) + 
  theme_minimal() +
  theme(legend.position="bottom")
  
```

The largest peak in the betting handle clearly coincides with the NFL regular season in September, and a smaller dip aligns with the season's end around the Super Bowl.

##### NCAA March Madness
```{r}
ncaa_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "03", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 3 ~ "Solid"
    ),
  
  label = case_when(
    month(date) == 3 ~ "March Madness"
    )
  ) |> 
  filter(date >= min(sb_rev_by_month$month), date <= max(sb_rev_by_month$month))


sb_rev_by_month |>
  mutate(handle_B = handle / 1e9) |>
  ggplot(aes(x = month, y = handle_B)) +
  geom_line() +
  ylab("Handle (in USD, Billions)") +
  geom_vline(
    data = ncaa_event_dates,
    aes(xintercept = date, linetype = type),
    color = "#009CDE",
    show.legend = TRUE 
  ) +
  scale_linetype_manual(
    name = "NCAA Events", 
    values = c("Solid" = "solid"), 
    labels = c("Solid" = "March Madness") 
  ) + 
  theme_minimal() +
  theme(legend.position="bottom")

```

A secondary peak in betting handle occurs during March Madness. This is important because it represents a period of intense, short-term betting interest.

### Cumulative Handle by State

```{r}
sb_rev_by_state = sb_rev_by_state |> 
  filter(market != "Total") |> 
  left_join(state_fips, join_by(market == state)) |> 
  left_join(
    select(state_legal_dates, state, first_start, online, offline), 
    join_by(market == state)) |> 
  mutate(
    legal_type = case_when(
      !is.na(online) & !is.na(offline) ~ "both",
      !is.na(offline) ~ "offline",
      !is.na(online) ~"online"
    )
  ) |> 
  mutate(
    hover = paste("<b>",market,"</b>", '<br>', 
                  "Handle: $", format(handle / 1000000000, digits = 2), "B", "<br>", 
                  "Revenue: $", format(revenue / 100000000, digits = 2), "M", "<br>",
                  "Hold: ", hold, '%<br>', 
                  "Taxes: ", format(taxes / 100000000, digits = 2), "M",
                  "<extra></extra>")
    )
                      
# give state boundaries a white border
l = list(color = toRGB("white"), width = 2)

# specify some map projection/options
g = list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig = plot_geo(sb_rev_by_state, locationmode = 'USA-states')
fig = fig |> 
  add_trace(
    z = ~handle, hovertemplate = ~hover, #text = ~hover, 
    locations = ~abbr,
    color = ~handle, colorscale="Viridis"
  ) |> 
  colorbar(title = "Millions USD") |> 
  layout(
    title = 'Cumulative Handle by State <br>(Hover for breakdown)',
    geo = g
  )

fig
```

### Handle by Channel in 2024

```{r eval=FALSE}
# ggplot version
sb_rev_by_state_month |> 
  mutate(handle_B = handle / 1e9) |>
  mutate(year = year(month)) |>   # Extract year
  group_by(state, year) |>        # Group by year
  summarize(
    total_handle = sum(handle_B, na.rm = TRUE),
  ) |> 
  filter(year == 2024) |> 
  ungroup() |> 
  mutate(
    state = fct_reorder(state, total_handle)
  ) |> 
  ggplot(aes(x=state, y= total_handle)) +
  coord_flip() +
  geom_bar(stat="identity") +
  labs(
    title = "Sports Betting Handle by State in 2024",
    x = "State",
    y = "Handle (in USD, Billions)"
  ) +
  theme_minimal()
```

Most sports betting occurs online, although there are a few states that only allow in person sports betting.

```{r}
#plotly version
sb_rev_by_state_month |> 
  mutate(year = year(month)) %>%  # Extract year
  group_by(state, year) %>%                   # Group by year
  summarize(
    total_handle = sum(handle, na.rm = TRUE),
  ) |> 
  filter(year == 2024) |> 
  ungroup() |> 
  mutate(state = case_when(
    state == "District of Columbia" ~ "DC",# Keep original row names for other categories
    TRUE ~ state
  )) |> 
  left_join(select(state_handle_channel, state, pct_online), by = "state") |> 
  mutate(
      state = fct_reorder(state, total_handle),
      online = if_else(!is.na(pct_online), total_handle * pct_online, 0),
      #online = total_handle * pct_online,
      in_person = if_else(!is.na(pct_online), total_handle * (1 - pct_online), 0),
      #retail = total_handle * (1 - pct_online),
      not_specified =  if_else(is.na(pct_online), total_handle, 0)
  ) |> 
  select(state, online, in_person, not_specified, pct_online) |> 
  pivot_longer(
    cols = c(online, in_person, not_specified),
    names_to = "channel"
  ) |> 
  mutate(
    hover = paste("<b>",state,"</b>", '<br>', 
                  "$", format(value / 1000000000, digits = 2), "B", "<br>", 
                  (pct_online * 100), "% online",
                  "<extra></extra>")) |> 
  plot_ly(y = ~state, x = ~value, color = ~channel, type = 'bar', hovertemplate = ~hover, orientation='h') |> 
  layout(
    barmode = 'stack',
    font = list(family = "Courier New", size = 8),
        #legend = list(orientation = 'v',
        #            yanchor = "center",
        #            traceorder = "reversed", y = 0.5,
        #            font = list(size = 10)),
         xaxis = list(title = list(text ='Cumulative Handle', font = list(size = 12))),
         yaxis = list(title = list(text ='', font = list(size = 12)))
    )
```

We can see New York has the highest handle which makes sense since it is the most populous state. We can get a sense of the popularity of sports gambling by looking at total handle per capita using US census data for total population by state.

### Handle per Capita in 2024

```{r eval=FALSE}
#ggplot version
sb_rev_by_state_month |> 
  mutate(year = year(month)) |>  
  group_by(state, year) |>                 
  summarize(
    total_handle = sum(handle, na.rm = TRUE),
  ) |> 
  filter(year == 2024) |> 
  ungroup() |> 
  left_join(select(state_pop, state, "2024"), by = "state") |> 
  rename(population_2024 = '2024') |> 
  mutate(
    handle_per_capita = total_handle / population_2024,
    state = fct_reorder(state, handle_per_capita)
  ) |> 
  ggplot(aes(x=state, y= handle_per_capita)) +
  coord_flip() +
  geom_bar(stat="identity") +
  labs(
    title = "Sports Betting Handle per Capita in 2024",
    x = "State",
    y = "Handle ($) per capita"
  ) +
  theme_minimal()
```

```{r}
sb_rev_by_state_month |> 
  mutate(year = year(month)) %>%  # Extract year
  group_by(state, year) %>%                   # Group by year
  summarize(
    total_handle = sum(handle, na.rm = TRUE),
  ) |> 
  filter(year == 2024) |> 
  ungroup() |> 
  left_join(select(state_pop, state, "2024"), by = "state") |> 
  rename(population_2024 = '2024') |> 
  mutate(
    handle_per_capita = total_handle / population_2024,
    state = case_when(
      state == "District of Columbia" ~ "DC",# Keep original row names for other categories
      TRUE ~ state
    ),
    state = fct_reorder(state, handle_per_capita)
  ) |> 
  plot_ly(y = ~state, x = ~handle_per_capita, type = 'bar', orientation='h') |> 
  layout(font = list(size = 8),
        legend = list(orientation = 'v',
                    yanchor = "center",
                    traceorder = "reversed", y = 0.5,
                    font = list(size = 10)),
         xaxis = list(title = list(text ='Handle per Capita ($)', font = list(size = 12))),
         yaxis = list(title = list(text ='', font = list(size = 12)))
         )
```

We can see that Nevada and New Jersey have the highest handle per capita.

### Interactive Handle by State and Sports season

Use the chart below to select states from the legend to compare to New York. You can use the buttons to toggle highlights on the NFL season (red with orange line for the Superbowl) or NCAA Basketball March Madness. 

```{r}
#Get revenue by state

states = sb_rev_by_state_month |> 
  select(state) |> 
  unique() |> 
  mutate(
    state_trace = str_replace_all(state, " ", "")
  ) |> 
  filter(
    state != "New York"
  )


state_hidden_df = sb_rev_by_state_month |> 
  select(month, state, handle) |> 
  mutate(
    state = str_replace_all(state, " ", "")
  ) |> 
  pivot_wider(
    names_from = state,
    values_from = handle
  ) |> 
  arrange(month) 
  #mutate(across(everything(), ~replace_na(., 0))) |> 

#create figure
fig = state_hidden_df |> 
  plot_ly(x = ~month, colors = "Set1") |> 
  layout(font = list(size = 8),
         xaxis = list(title = list(text ='Date', font = list(size = 12))),
         yaxis = list(title = list(text ='Handle ($)', font = list(size = 12)))
         ) |> 
  add_trace(y = ~NewYork, name = "New York",  type = 'scatter', mode = 'lines', visible = TRUE, color="N")

#add in lines for NFL season and march madness
years_in_data = unique(year(sb_rev_by_state_month$month)) 

nfl_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "09", "01", sep = "-")),
    ymd(paste(years_in_data, "01", "01", sep = "-")),
    ymd(paste(years_in_data, "02", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 2 ~ "Solid", 
    TRUE ~ "Dotted" 
    ),
  
  label = case_when(
    month(date) == 9 ~ "Start of Season",
    month(date) == 1 ~ "End of Season",
    month(date) == 2 ~ "Super Bowl"
    )
  ) |> 
  filter(date >= min(sb_rev_by_state_month$month), date <= max(sb_rev_by_state_month$month))

nfl_season = nfl_event_dates |> 
  filter(type == "Dotted") |> 
  arrange(date) |> 
  mutate(
    year = year(date - months(6))
  ) |> 
  pivot_wider(
    names_from = "label",
    values_from = "date"
  ) |> 
  janitor::clean_names()

max_y = max(pull(sb_rev_by_state_month, handle))

nfl_dotted = pull(filter(nfl_event_dates, type=="Dotted"), date)
nfl_solid = pull(filter(nfl_event_dates, type=="Solid"), date)

ncaa_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "03", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 3 ~ "Solid"
    ),
  
  label = case_when(
    month(date) == 3 ~ "March Madness"
    )
  ) |> 
  filter(date >= min(sb_rev_by_state_month$month), date <= max(sb_rev_by_state_month$month))

ncaa_solid = pull(ncaa_event_dates, date)

for (i in 1:nrow(states)) {
  row = states[i,]
  fig = fig |> add_trace(
    y =as.formula(paste0("~", row["state_trace"])), 
    name=row["state"],
    type = 'scatter', 
    mode = 'lines', 
    visible = 'legendonly',
    color = paste(substr(row["state_trace"], 1, 1), substr(row["state_trace"], nchar(row["state_trace"]), nchar(row["state_trace"])))
    )
}

nfl_shape_list = list()

for (i in 1:nrow(nfl_season)) {
      start_d <- nfl_season$start_of_season[i]  # Access 'Name' column for the current row
      end_d <- nfl_season$end_of_season[i]    # Access 'Age' column for the current row

      print(paste("Row", i, ":", start_d, "is", end_d, "years old and lives in"))
      
      year_shape = list(
          type = "rect",
          x0 = start_d,
          x1 = end_d,
          y0 = 0,
          y1 = max_y,
          yref = "paper", # This ensures the line spans the entire y-axis height
          fillcolor = "red", line = list(color = "red"), opacity = 0.3
          #line = list(dash="dot", color = "red", width = 1)
          #visible = FALSE, # Initially hidden
        )
      nfl_shape_list[[i]] = year_shape
}

nfl_shape_list = c(nfl_shape_list,
                   lapply(nfl_solid, function(date_val) {
        list(
          type = "line",
          x0 = date_val,
          x1 = date_val,
          y0 = 0,
          y1 = max_y,
          yref = "paper", # This ensures the line spans the entire y-axis height
          line = list(color = "orange", width = 0.7, dash = "dot")
        )
      }))

ncaa_shape_list = lapply(ncaa_solid, function(date_val) {
        list(
          type = "line",
          x0 = date_val,
          x1 = date_val,
          y0 = 0,
          y1 = max_y,
          yref = "paper", # This ensures the line spans the entire y-axis height
          line = list(dash="dot", color = "blue", width = 1)
        )
      })

shapes = c(lapply(nfl_dotted, function(date_val) {
        list(
          type = "line",
          x0 = date_val,
          x1 = date_val,
          y0 = 0,
          y1 = max_y,
          yref = "paper", # This ensures the line spans the entire y-axis height
          line = list(dash="dot", color = "red", width = 1)
          #visible = FALSE, # Initially hidden
        )
      }),
    lapply(nfl_solid, function(date_val) {
        list(
          type = "line",
          x0 = date_val,
          x1 = date_val,
          y0 = 0,
          y1 = max_y,
          yref = "paper", # This ensures the line spans the entire y-axis height
          line = list(color = "orange", width = 0.5, dash = "dot")
          #visible = FALSE, # Initially hidden
        )
      }))


#add buttons to show/hide shapes
fig = fig |>  layout(updatemenus = list(
    list(
      type = "buttons",
      direction = "right",
      xanchor = "left",
      yanchor = "top",
      x = 0.1, # Position the menu at 10% from the left
      y = 1.2, # Position the menu at 110% from the bottom
      pad = list("r" = 10, "t" = 10),
      buttons = list(
        list(
          method = "relayout",
          args = list(list(shapes = list())),
          label = "Hide seasons"
        ),
        list(
          method = "relayout",
          label = "Show NFL season",
          args = list(list(shapes = nfl_shape_list))
        ),
        list(
          method = "relayout",
          label = "Show NCAA season",
          args = list(list(shapes = ncaa_shape_list))
        )
      )
    )
  ))

fig
```


## BRFSS

### Data Completeness

One of our primary concerns with data was ensuring that the key outcome variables were reported consistently from 2017 to 2024.

```{r}
all_outcome_vars = c("general_health", "general_health_refactored", "michd",
                 "physical_health", "physical_health_not_good_days", "leisure_physical_activity_last_30_days",
                 "mental_health", "mental_health_not_good_days", "poor_health", "depressive_disorder",
                 "difficulty_self_care", "life_satisfaction", "emotional_support", "loneliness", "binge_drink",
                 "heavy_drink", "lost_reduced_employment", "financial_strain_bills", "financial_strain_utilities")

summarize_outcome_by_year = function(outcome_var, df = brfss_data) {
  
  outcome_sym = rlang::sym(outcome_var) 
  
  df |> 
    group_by(year(date), !!outcome_sym) |> 
    summarize(n = n(), .groups = "drop") |> 
    pivot_wider(
      names_from = `year(date)`, 
      values_from = n
    )
}

find_complete_years = function(outcome_var, summary_df){
  
  outcome_sym = rlang::sym(outcome_var) 
  
  summary_df |> 
    filter(!is.na(!!outcome_sym)) |> 
    summarize(
      across(
        .cols = everything(),
        .fns = ~ !all(is.na(.x)) 
        )
      )|>
    select(-!!outcome_sym) |> 
    pivot_longer(
      cols = everything(),
      names_to = "year",
      values_to = "is_complete"
      ) |> 
    summarize(
      n_complete_years = sum(is_complete)
      ) |> 
    pull(n_complete_years)
    
}

complete_outcomes = tibble(
  outcome = all_outcome_vars,
  summary = map(all_outcome_vars, summarize_outcome_by_year),
  n_complete_years = map2(all_outcome_vars, summary, find_complete_years)
  ) |> 
  arrange(n_complete_years) |> 
  filter(
    n_complete_years == 8
  ) |> 
  pull(outcome)

setdiff(all_outcome_vars, complete_outcomes)
```

Some of the outcome variables we were interested in looking at like financial strain, were not available in all years. In order to maintain a robust analysis over time, we were limited to those variables that were available over all years. Our 4 primary outcomes are available over all years. 

### Distribution of Key Outcomes + Rationale for Binary Outcomes

```{r}
brfss_data |>
  filter(year(date) %in% c(2017, 2024)) |>
  select(year = date, mental_health, physical_health) |>
  mutate(
    year = factor(year(year)),
    mental_health = ifelse(mental_health > 30, 31, mental_health), 
    physical_health = ifelse(physical_health > 30, 31, physical_health)
  ) |>
  pivot_longer(
    cols = c(mental_health, physical_health),
    names_to = "outcome",
    values_to = "n_bad_days"
  ) |>
  ggplot(aes(x = n_bad_days, fill = year)) +
  geom_bar(position = "dodge") +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 31), labels = c(0, 7, 14, 21, 31)) +
  scale_fill_manual(values = c("2017" = "cadetblue3", "2024" = "darkslateblue")) +
  facet_wrap(~outcome) +
  labs(
    title = "Distribution of Poor Health Days (2017 vs. 2024)",
    x = "Number of Days of Poor Health (Past 30 Days)",
    y = "Count of Respondents"
  ) +
  
  theme_minimal() +
  theme(legend.position="bottom")
```

Given the concentration of responses at 0 days, we decided to use a binary outcome variable like `any_physical_health_not_good_days` as the outcome variable. This also follows what Couture, et. al. used in their analysis. It also simplified statistical tests that we decided to do later on.


### Correlation Analysis

To understand relationships among the health outcome variables, we conducted a correlation analysis. Initially, we computed separate correlation matrices for males and females to assess whether the relationships among health outcomes differed meaningfully by sex. However, this stratification did not reveal significant differences, therefore, we analyzed correlations using the full sample.

```{r}
#selecting variables
correlation_vars = c("general_health","general_health_refactored", 
                     "mental_health", "mental_health_not_good_days",
                     "physical_health", "physical_health_not_good_days",
                     "poor_health", "depressive_disorder",
                     "difficulty_self_care", "michd",
                     "binge_drink", "heavy_drink", 
                     "leisure_physical_activity_last_30_days")


brfss_all = brfss_data |> 
  select(all_of(correlation_vars)) |> 
  mutate_all(as.numeric)

cor(brfss_all, 
      method = "spearman",
      use = "pairwise.complete.obs") |> 
  ggcorrplot(
    type = "upper",
    method = "square",
    lab = TRUE,
    lab_size = 2.5,
    tl.cex = 10,
    title = "Correlations"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

The correlation matrix revealed redundancy with pairs of variables, e.g., `mental_health` vs. `mental_health_not_good_days` (r = 0.99). Including both variables in regression models would introduce multicollinearity, so one variable from each pair was selected.

We can see that `depressive_disorder` showed the strongest positive correlation with `mental_health`(r = 0.44). `poor_health` showed a similar pattern correlating the strongest with `physical_health` (r = 0.43). These trends may reflect how individuals reporting depressive symptoms or more days where poor health negatively impacted usual activities also tend to report more poor physical and mental health days.

On the other hand, `heavy_drink` and `binge_drink` showed weak correlations with most other health outcomes.

Note: BRFSS contains a variable `general_health` which is a Likert scale from 1-6 with 1 being “Good Health” and 6 being “Poor Health.” There is also a `general_health_refactored` column with 1 being “Good Health” and 2 being “Poor Health.” `general_health_refactored` shows negative correlations with variables like `physical_health` and `mental_health`. As general health gets better (lower score), the number of bad health days goes up. This is an unexpected result. We can see that `general_health` has little correlation between most other health outcomes, which was surprising. We may be seeing these trends since `general_health_refactored` has fewer, more broadly defined categories. The more narrowly defined `general_health` shows minimal correlation. If we were to do future analyses, we would want to use the narrowly defined variable `general_health.`


### Distribution of Primary Health Outcomes by Demographic Variables

We explored how our primary health outcomes were distributed across demographic and socioeconomic groups. We used `mental_health_not_good_days`, `physical_health_not_good_days`, `depressive_disorder`, and `binge_drink` based on the consistent availability of these health outcomes across 2017-2024.

We used stacked bar plots to visualize these distributions with each bar representing a demographic category, and the bar divided into segments to show the proportion of individuals falling into each health outcome.

To efficiently visualize the distribution of these health outcome across any demographic variable, we constructed a set of plotting functions for each outcome. Here are some notable findings:


```{r}
#Bad mental health days outcome
plot_mental_health = function(demo_var, plot_title) {
  
    brfss_data |> 
    filter(
      !is.na(.data[[demo_var]]),
      !is.na(mental_health_not_good_days)
    ) |> 
  
  ggplot(aes(x = .data[[demo_var]], fill = mental_health_not_good_days)) +
    geom_bar(position = "fill") +
    scale_fill_manual(
    values = c(
      "0" = "darkolivegreen2",
      "1-13" = "yellow2",
      "14+" = "brown2"
    ),
    name = "# of Bad Mental Health Days"
  ) +
    labs(
      x = demo_var,
      y = "Proportion",
      title = plot_title
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
}

#Bad physical health days outcome
plot_physical_health = function(brfss_data, demo_var, plot_title) {
  
    brfss_data |> 
    drop_na(physical_health_not_good_days) |> 
    drop_na(.data[[demo_var]]) |> 
  
  ggplot(aes(x = .data[[demo_var]], fill = physical_health_not_good_days)) +
    geom_bar(position = "fill") +
    scale_fill_manual(
    values = c(
      "0" = "darkolivegreen2",
      "1-13" = "yellow2",
      "14+" = "brown2"
    ),
    name = "# of Bad Physical Health Days"
  ) +
    labs(
      x = demo_var,
      y = "Proportion",
      title = plot_title
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
}

#Depressive disorder outcome
plot_depression = function(demo_var, plot_title) {
  
    brfss_data |> 
    mutate(
      income_level = factor(
        income_level,
        levels = c(
          "Less than $10,000",
          "$10,000 to < $15,000",
          "$15,000 to < $20,000",
          "$20,000 to < $25,000",
          "$25,000 to < $35,000",
          "$35,000 to < $50,000",
          "$50,000 to < $75,000",
          "$75,000 or more")
        )
      ) |> 
    drop_na(depressive_disorder) |> 
    drop_na(.data[[demo_var]]) |> 
  
  ggplot(aes(x = .data[[demo_var]], fill = depressive_disorder)) +
    geom_bar(position = "fill") +
    scale_fill_manual(
    values = c(
      "No" = "steelblue2",
      "Yes" = "red4"
    ),
    name = "Depressive Disorder"
  ) +
    labs(
      x = demo_var,
      y = "Proportion",
      title = plot_title
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

#Binge drinker outcome
plot_bingedrink = function(demo_var, plot_title) {
  
    brfss_data |> 
    drop_na(binge_drink) |> 
    drop_na(.data[[demo_var]]) |> 
  
  ggplot(aes(x = .data[[demo_var]], fill = binge_drink)) +
    geom_bar(position = "fill") +
    scale_fill_manual(
    values = c(
      "No" = "steelblue2",
      "Yes" = "red4"
    ),
    name = "Binge Drinker"
  ) +
    labs(
      x = demo_var,
      y = "Proportion",
      title = plot_title
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```


#### Distribution of Bad Mental Health Days by Age Group:

Below are some selected plots, but you can view the complete demographic distributions on our [Shiny Dashboard](**ADD LINK HERE**).

```{r}
brfss_data |> 
  filter(
    !is.na(age_group_5yr),
    !is.na(mental_health_not_good_days)
  ) |> 
  
  ggplot(aes(x = age_group_5yr, fill = mental_health_not_good_days)) +
  geom_bar(position = "fill") +
  scale_fill_manual(
    values = c(
      "0" = "darkolivegreen2",
      "1-13" = "yellow2",
      "14+" = "brown2"
    ),
    name = "# of Bad Mental Health Days"
  ) +
  labs(
    x = "Age Group",
    y = "Proportion",
    title = "Distribution of Bad Mental Health Days by Age Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

The 18-24 and 25-29 age groups have the largest proportion of individuals reporting 1-13 and 14+ bad mental health days. Younger adults report the highest frequency of bad mental health days, while older adults report the lowest.

#### Distribution of Bad Physical Health Days by Sex


```{r}
brfss_data |> 
    drop_na(physical_health_not_good_days) |> 
    drop_na(sex) |> 
  
  ggplot(aes(x = sex, fill = physical_health_not_good_days)) +
    geom_bar(position = "fill") +
    scale_fill_manual(
    values = c(
      "0" = "darkolivegreen2",
      "1-13" = "yellow2",
      "14+" = "brown2"
    ),
    name = "# of Bad Physical Health Days"
  ) +
    labs(
      x = "Sex",
      y = "Proportion",
      title = "Distribution of Bad Physical Health Days By Sex"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

We can see that overall distributions are very similar across males and females. The proportion reporting 14+ days is relatively small and nearly identical between sexes. The similarity in patterns can show that sex differences are unlikely to significantly confound associations unless interacting with age or other factors.


#### Distribution of Depressive Order by Income Level

```{r}
brfss_data |> 
  mutate(
    income_level = factor(
      income_level,
      levels = c(
        "Less than $10,000",
        "$10,000 to < $15,000",
        "$15,000 to < $20,000",
        "$20,000 to < $25,000",
        "$25,000 to < $35,000",
        "$35,000 to < $50,000",
        "$50,000 to < $75,000",
        "$75,000 or more")
    )
  ) |> 
  drop_na(depressive_disorder) |> 
  drop_na(income_level) |> 
  
  ggplot(aes(x = income_level, fill = depressive_disorder)) +
  geom_bar(position = "fill") +
  scale_fill_manual(
    values = c(
      "No" = "steelblue2",
      "Yes" = "red4"
    ),
    name = "Depressive Disorder"
  ) +
  labs(
    x = "Income Level",
    y = "Proportion",
    title = "Distribution of Depressive Order by Income Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
```

The lowest-income groups (less than \$10,000 and \$10,000 to \<\$15,000) have the highest proportions of individuals reporting a depressive disorder. As income decreases, the proportion of respondents with depressive disorder increases.


#### Distribution of Binge drinking by Age Group

```{r}
brfss_data |> 
  drop_na(binge_drink) |> 
  drop_na(age_group_5yr) |> 
  
  ggplot(aes(x = age_group_5yr, fill = binge_drink)) +
  geom_bar(position = "fill") +
  scale_fill_manual(
    values = c(
      "No" = "steelblue2",
      "Yes" = "red4"
    ),
    name = "Binge Drinker"
  ) +
  labs(
    x = "Age Group",
    y = "Proportion",
    title = "Distribution of Binge Drinkers by Age Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We can see that binge drinking is most common among younger adults with the peak in the 25-29 age group. It steadily declines as age increases.





